The abundance of Bullet-Groups in $\Lambda$CDM

### 2014-03-20 paper/ms.tex
keywords
goes for a second round of comments

\documentclass{emulateapj}
%\documentclass{aastex}
\submitted{{\it Submitted for publication in ApJL}}
\usepackage{multirow,color,wrapfig,ulem}
\usepackage {graphicx}

@@

We estimate the expected distribution of displacements between the two
dominant dark matter density peaks and between baryons and dark matter
in halos simulated within a full cosmological context in $\Lambda$CDM.
We use as a benchmark the observation of \bullg, which is the lowest mass system observed so far
featuring a bi-modal dark matter distribution with a dislocated
baryonic component. We extend to lower masses previous computational results by
studying two halo samples: groups and clusters. We find that $(50 \pm
10)\%$ of the dark matter halos with circular velocities in the range
$300\kms$ to $700\kms$ (groups) show multimodal morphologies with
displacements between the dark matter clumps equal or larger than
$186 \pm 30$\hkpc as observed in \bullg. For dark matter halos with
circular velocities larger than $700\kms$ (clusters) this fraction
rises to $(70 \pm 10)\%$. Using the same simulation we estimate the
dark matter-to-baryon spatial separation and find that $0.1$ to
$1.0\%$ of the groups should present separations equal or
larger than $87\pm 14$\hkpc corresponding to our observational
benchmark; for clusters this fraction rises to $(7\pm 3)\%$, consistent
with previous studies of dark matter to baryon separations. In terms
of absolute number densities we find that the expected value for
groups similar to \bullg\ is three times larger than the expected value
for clusters. This extension of the theoretical predictions and
observational results towards low mass bi-modal dark matter
configurations opens up the possibility for a new statistical test of
$\Lambda$CDM.   
\end{abstract}

\keywords{dark matter, galaxies: clusters: individual (SL2S
  J08544-0121), galaxies: interactions, methods: numerical}

@@

Groups around this mass are $\sim10$ times more abundant than clusters
in the mass range of the Bullet Cluster $>10^{15}$\hMsun. This 
should open up the possibility of finding Bullet Groups in large
numbers to test $\Lambda$CDM. However, a larger abundance of small
mass systems has to be weighted by the probability of being in a
merger and  presenting a large displacement between the DM and
baryonic components.  These two conditions (merger rates, maximum
possible displacement) are a funcion of DM halo mass in $\Lambda$CDM
cosmologies.  A detailed statistical study to estimate the abundance
of Bullets has been performed for clusters but not for lower mass
systems \citep{ForeroRomero2010}.

@@

The angle between these two vectors can be quantified by, 
\begin{equation}
  \mu\equiv
  \cos(\theta)=\frac{\vec{v{}}\cdotp{}\vec{r}}{\left\|\vec{v}{}\right\|
    \left\|\vec{r}\right\|} .
 \end{equation} 
%
This encondes information about the collision, i.e. cases of
$|\mu|\approx 1$ can be considered as head-on collisions while
$|\mu|\approx 0$ describes a grazing trajectory.

@@

We also use the physical quantities described above to discriminate
between grazing trajectories, $|\mu|<0.9$, and Bullet-like collissions
with $|\mu|\approx 1$. In the latter case we can also distinguish three
main stages in a such encounter. First, when the sub-halo crosses the
virial radius of the host halo starting a head on collision, $D_{\rm
  off} \approx 1$ and $\mu\approx -1$. Second, as the sub-halo crosses
for the first time the center of the host halo $D_{\rm off}<1.0$ and
$\mu\approx 1$. Third, as the sub-halo reaches apogee and comes back
to the center of the halo $D_{\rm off} < 1.0$ and $\mu< -1$.

@@

Figure \ref{fig:displacement} presents the integrated
probability distribution for the DM-DM displacements, $d_{\rm 2D}$ The
left panel shows the displacement in physical units and the right
panel as a fraction of the virial radius of the host halo.  The panel
with the projected 2D physical displacements also shows a vertical
stripe with the estimated displacement for the Bullet-group reported
by \cite{Gastaldello}.

@@

\subsection{DM-baryon Displacement}
\label{sec:baryonic_displacements}

@@

The results for the integrated distributions for $>d_{\rm 2D}^{\rm bar}$
are shown in Figure \ref{fig:baryonic_displacements}. The dashed lines
represented the results for groups and the continuous lines correspond
to clusters. As a test of our model we compare the cluster results
against the analytic fit provided by \cite{ForeroRomero2010}. This fit
reproduces the statistics for the DM-baryon separation found for
clusters more massive than $>10^{14}$\hMsun\ in a simulation
which included a description for DM and gas with $8$ times more volume
and $8$ times less mass resolution than the Bolshoi Simulation. The
fit is valid for separations larger than $70$\hkpc, beyond which we
find that it provides a remarkably good description within a factor of
$\sim 2$ of our results. This gives us confidence in our approach to
estimate the expected fraction of groups with a DM-baryon
displacement.

>>>>

\documentclass{emulateapj}
%\documentclass{aastex}
\submitted{{\it Submitted for publication in ApJL}}
\usepackage {graphicx}

@@

We estimate the expected distribution of displacements between the two
dominant dark matter (DM) density peaks (DM-DM displacements) and between
gaseous baryons and dark matter (DM-baryon displacements) in DM halos 
with masses larger than $10^{13}$\hMsun. We use as a benchmark the
observation of \bullg, which is the lowest mass system ($1.0\times
10^{14}$\hMsun) observed so far
featuring a bi-modal dark matter distribution with a dislocated
baryonic component. We find that $(50 \pm 10)\%$ of the dark matter
halos with circular velocities in the range $300\kms$ to $700\kms$
(groups) show DM-DM displacements equal or larger than $186 \pm
30$\hkpc as observed in \bullg. For dark matter halos with circular
velocities larger than $700\kms$ (clusters) this fraction rises to
$(70 \pm 10)\%$. Using the same simulation we estimate the DM-baryon
displacements and find that $0.1$ to $1.0\%$ of the groups should
present separations equal or larger than $87\pm 14$\hkpc corresponding
to our observational benchmark; for clusters this fraction rises to
$(7\pm 3)\%$, consistent with previous studies of dark matter to
baryon separations. Considering both constraints on the DM-DM and
DM-baryon displacements we find that the number density of groups
similar to \bullg\ is $\sim 6.0\times 10^{-7}$ Mpc$^{-3}$, three times
larger than the estimated value forclusters. These results open up
the possibility for a new statistical test of $\Lambda$CDM.     
\end{abstract}

\keywords{dark matter --- galaxies: clusters: individual (SL2S
  J08544-0121) --- galaxies: interactions --- methods: numerical}

@@

Groups around this mass are $\sim10$ times more abundant than clusters
in the mass range of the Bullet Cluster $>10^{15}$\hMsun. This 
should open up the possibility of finding Bullet Groups in large
numbers to test $\Lambda$CDM. However, a larger abundance of small
mass systems has to be weighted by the probability of being in a
merger and  presenting a large displacement between the DM and
baryonic components.  These two conditions (merger rates, maximum
possible displacement) are a funcion of DM halo mass in $\Lambda$CDM
cosmologies.  A detailed statistical study to estimate the DM-gas
displacements of Bullets has been performed for clusters
\citep{ForeroRomero2010} but not for lower mass systems.

@@

The angle between these two vectors can be quantified by, 
\begin{equation}
  \mu\equiv
  \cos(\theta)=\frac{\vec{v{}}\cdotp{}\vec{r}}{\left\|\vec{v}{}\right\|
    \left\|\vec{r}\right\|} .
 \end{equation} 
%
This encondes information about the collision, i.e. cases of
$|\mu|\sim 1$ can be considered as head-on collisions while
$|\mu|< 0.9$ describes a grazing trajectory.

@@

We also use the physical quantities described above to discriminate
three main stages in a Bullet-like encounter with $|\mu|\sim
1$. First, when the sub-halo crosses the virial radius of the host
halo starting a head on collision, $D_{\rm   off} \sim 1$ and $\mu\sim
-1$. Second, as the sub-halo crosses for the first time the center of
the host halo $D_{\rm off}<1.0$ and $\mu\sim 1$. Third, as the
sub-halo reaches apogee and comes back to the center of the halo
$D_{\rm off} < 1.0$ and $\mu\sim -1$.

@@

Figure \ref{fig:displacement} presents the integrated
probability distribution for the DM-DM displacements, $d_{\rm 2D}$. The
left panel shows the displacement in physical units and the right
panel as a fraction of the virial radius of the host halo.  The panel
with the projected 2D physical displacements also shows a vertical
stripe with the estimated displacement for the Bullet-group reported
by \cite{Gastaldello}.

@@

\subsection{DM-baryon Displacements}
\label{sec:baryonic_displacements}

@@

The results for the integrated distributions for $>d_{\rm 2D}^{\rm bar}$
are shown in Figure \ref{fig:baryonic_displacements}. The dashed lines
represented the results for groups and the continuous lines correspond
to clusters. As a test of our model we compare the cluster results
against the analytic fit provided by \cite{ForeroRomero2010}. This fit
reproduces the statistics for the DM-baryon separation found for
clusters more massive than $>10^{14}$\hMsun\ in a simulation
which included a description for DM and gas with $8$ times more volume
and $8$ times less mass resolution than the Bolshoi Simulation. The
fit is valid for separations larger than $70$\hkpc, beyond which we
find that it provides a remarkably good description within a factor of
$\sim 2$ of our results. This gives us confidence in our approach to
correctly estimate the expected fraction of groups with a DM-baryon
displacement.
